Stage-Splitting a Modal Language

### 2014-09-29 paper/intro.tex
moving around some stuff; merging techreport into paper

We often find ourselves in situations where the input to a computation arrives in two parts, at different points in time.

The natural response is to try and do some part of the computation early, and the rest of the computation once the second input arrives.

Since the computation runs in two parts, the total cost can be split as well, into $m+n$.

Additionally, we can run the second stage multiple times.  Cost is now $m+bn$.

Jorring et al. (\cite{jorring86}) identify three classes of staging techniques: meta-programming, partial evaluation, and stage-splitting.  The first two of these have received significantly more attention than the third.  Countless meta-programming systems exist (...twelve thousand citations...\cite{devito13}), and their background theory and type-systems are well understood (\cite{davies01}).  Partial evaluation, too, is well-understood.  Partial evaluation systems exist... . This paper explores both theory and applications for stage-splitting.

>>>>

It is a common observation that some computations can be performed efficiently by using {\em staging}, wherein the execution of a program is split over two or more discrete points in time ({\em i.e.} stages).  This most often arises when we want to evaluate a multivariate function several times, with some inputs dynamic (varying across invocations) and some static (constant across invocations).  Examples include,
\begin{itemize}
\item Computing $b^e$, where $b$ varies and $e$ is constant.
\item Finding the $i$th largest element in a list, where the list is constant and $i$ varies.
\item Computing the shading of a surface in a 3D rendering, where some contributors to the final pixel color can be approximated as constant over  pixel blocks.  
\end{itemize}

With these goals in mind, our initial concern is how to express multistage programs. To this end, we have created \lang, a typed multi-stage lambda calculus featuring products, sums, and isorecursive types. An important property of \lang\ is that the staging of a well-typed program is apparent without resorting to any variable analysis, contra standard binding time analysis.

Once we have appropriately written {\em multi-stage} programs, we then must execute them in a way that respects the staging.  In this paper, we explore two techniques for this: {\em partial evaluation} and {\em stage-splitting}.

The main idea of the former technique is to evaluate all of the constant portions of the program and embed the partial results into a {\em residual} program that depends only on the varying inputs.  This is an old and well-explored idea, though the typical setup of partial evaluation implies a more powerful binding time analysis than is necessary for our language.

The main idea of the latter technique is to statically compile the multi-stage program into several smaller programs, one per stage, and then to execute those in a standard way once inputs are available.  This has previously been explored as an optimization technique for languages with fewer features (notably, no functions) or a more constrained staging model than \lang. Prior examples include the ``data specialization" work of Knoblock and Ruff\,\cite{knoblock96} as well as the compilation of high-level shading languages for computer graphics such as RTSL\,\cite{Proudfoot:2001} and Spark\,\cite{Foley:2011}.  Unlike partial evaluation, stage splitting is purely a static program transformation and is not dependent on the input data.

\TODO motivate this example

\Cref{sec:semantics} describes the grammar and type system of \lang, as well as a reference semantics which is essentially a partial evaluator that respects staging. \Cref{sec:splitting} describes a stage-splitting transformation for \lang\ and proves its equivalence to the reference semantics.
Finally, \Cref{sec:examples} gives several examples of how stage-splitting terms in \lang\ can be used to derive classic data structures, and discusses the effectiveness and limitations of this technique.  Notably, these examples make use of the stage-splitting features not available in prior work.
